Attrition in Online Learning: Understanding Persistence and Dropout in Massive Open Online Courses

### 2014-10-28 paper_v1.Rnw
made edits based on emily's comments

Educational environments have become increasingly diverse. Traditional schools and universities have a characteristically rigid structure, including instructor-defined---even nationally agreed---syllabi, fixed time schedules, entry requirements, and material costs to enter and exit. Novel institutional structures have been developed to overcome particular constraints. Community colleges, for instance, were created in an attempt to democratize education by offering instruction at a lower cost and by accommodating people with less flexible schedules \cite{goldrick2010challenges}. Distance learning programs intended to deliver education in remote parts of the world and for people who simply could not attend in-person classes. Course materials, including assessments, were delivered through mail (``correspondence education''), radio, television, and eventually the Internet, thereby addressing geographical and time related constraints of traditional instruction \cite{moore1996distance}.

The latest generation of online learning environments, characterized by massive open online courses (MOOCs), has pushed the boundary on the scale of education \cite{waldrop2013campus}. By design, MOOCs provide course materials to millions of people worldwide. This scale could be achieved by pre-recording lectures, designing assessments that can be graded automatically, and by leveraging the momentum of the number of people involved (e.g., to facilitate peer learning or peer grading \cite{kulkarni2013peer,cambre2014talkabout}). Maybe by virtue of their large scale, their prominent instructors, or their adherence to contemporary interface designs, MOOCs rapidly became an online media phenomenon. People would sign up weeks in advance of the course launch date, many of whom would never even enter the course site. And among those who enroll and enter the site, a large proportion tends to only ``sample'' some content and leave again \cite{kizilcec2013deconstructing}. Many of the prototypical behaviors observed in MOOCs \cite{kizilcec2013deconstructing,breslow2013studying} (other?) resemble those on online media platforms, such as YouTube or tumblr. This trend has also been reflected in the diversity of MOOC learners' motivations for enrolling \cite{kizilcec2015motivation}.

Shortly after the first wave of courses had finished, extensive media coverage led to MOOCs becoming associated with high attrition rates \cite{lewin2013after,parr2013mooc,guthrie2013moocs}. Early MOOC research cautioned against dichotomizing learners into successes and failures based on course completion \cite{kizilcec2013deconstructing,rivard2013measuring}. Instead, more nuanced categorizations based on learner behavior \cite{kizilcec2013deconstructing,clow2013moocs} (other?), motivations \cite{kizilcec2015motivation}, or intentions \cite{wilkowski2014student} have been suggested. Ultimately, perspectives on persistence and attrition in MOOCs depend on how MOOCs have been conceptualized. Kizilcec and Schneider \citeyear{kizilcec2015motivation} proposed that MOOCs have bridged two different world: one world is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other world adheres to the ``grammar of schooling'', which presupposes instructor-defined goals that students strive to achieve \cite{tyack1994grammar}. Viewing MOOC participation as bridging these two worlds undoubtedly adds a layer of complexity to interpretations of attrition.

This paper presents a systematic investigation into attrition in MOOCs, based on self-report and behavioral data collected from 20,879 online learners in 18 courses. A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition and thereby enables targeted interventions to help learners who want to persist. We begin by briefly reviewing the large literature on attrition in educational environments with a focus on important developments in understanding its causes. Building on this foundation of prior work, Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions. In Study 2, to develop a deeper understanding of MOOC attrition, we invited around 6,500 learners who were predicted to have disengaged from a MOOC to provide feedback. Their open response answers about what challenges they faced while taking the MOOC were iteratively coded. We investigated the role of volitional control for learners who report having not enough time, and examine goal striving, social belonging, and growth mindset in relation to being successful in a MOOC.

@@

The majority of early work on attrition centered around theoretical models of students' decision to persist or dropout of a traditional higher education setting. An early model suggested that students' presistence is largely driven by their prior behavior, attitudes, and norms \cite{fishbein1975belief}. The psychological processes involved in turning an intent to learn into the decision to persist were thought to be mediated by volition, i.e., the extent to which the student engages in goal-direct behaviors in the face of distraction \cite{corno1993role}. Hence, motivation alone is necessary but not sufficient for persistence. Students may fail to sustain efforts in the absence of strong self-regulatory skills.

\begin{description}
  \item[RQ1] How do prior learner characteristics (demographics, geographic location, intentions and motivations, and previous online courses) influence learners' likelihood of dropping out of a MOOC?
  \item[H1] Successful MOOC learners exhibit higher levels of goal striving than unsuccessful ones.
\end{description}

@@

\begin{description}
  \item[RQ2] What proportion of learners who dropped out were satisfied with their achievements in the course?
  \item[H3] Successful MOOC learners have a more fluid conception of intelligence (i.e., more of a growth than fixed mindset) than unsuccessful ones.
\end{description}

@@

\begin{description}
\item[RQ3a] What reasons do learners report for dropping out of a MOOC?
\item[RQ3b] How do reasons for dropping out vary across prior learner characteristics (cf. RQ1)?
\item[RQ4] How does behavior in MOOCs vary by reasons for dropping out?
\end{description}

@@

(Note: this section is work in progress, we will go through each RQ and present results that address them here.)

@@

The two most commonly reported reasons for dropout are related to one's ability to commit time to the course. The correlation between the two reasons is $r=0.32$. A more detailed analysis we conditioned on responses to these two reasons revealed that 30\% of respondents who reported at least one of these two time reasons reported no other reasons for dropout, while 24\% reported one other reason and 46\% reported more than one.

Difficulty of an assignment or exam was the third most reported reason for dropout. It is most positively correlated with reasons citing difficulties, including ``Too advanced for my background'' ($r=0.4$), ``Materials not explained properly'' ($r=0.37$), and ``Technical difficulties'' ($r=0.33$).

The proportion of respondents who selected ``Technical difficulties'' exceeded 20\% in 10 out of the 17 courses. This reason, however, exhibits positive correlations stronger that 0.2 with 7 other reasons. [What might we like to study more about respondents who selected this reason?]

In the median, 19\% of respondents reported they learned what they wanted in the course, and this reason was moderately positively correlated with reasons citing no or incomplete interest in the course, namely ``I was only exploring the course'' ($r=0.36$), ``The course is not challenging enough for me'' ($r=0.31$), and ``The course is not what I was looking for'' ($r=0.28$).

Another set of interesting positive correlations exists between ``Course is not what I was looking for'' and other reasons citing the materials' level of difficulty/challenge, the value of the credentials to the respondent, and the extent to which she enjoyed the online course format, suggesting .... [TBC]

The reason that exhibited the strongest positive correlations with other features was ``Videos were confusing''. it correlates most strongly with ``Materials not explained properly'' ($r=0.66$) and ``Did not enjoy online course format'' ($r=0.54$)

Unlike correlations among reasons for disengagement, correlations of reasons for disengagement with enrollment intentions were mostly weak ($-0.14 < r < 0.24$). Among the strongest positive correlations were the cases listed in Table \ref{tab:rfd_ei_strongest_correlations}

@@

\section{Acknowledgments}
% We thank Elise Ogle and Ruth Bram for helping with the development of the codebook. 
Omitted for blind review.

>>>>

Educational environments have become increasingly diverse. Traditional schools and universities have a characteristically rigid structure, including instructor-defined---even nationally agreed---syllabi, fixed time schedules, entry requirements, and material costs to enter and exit. Novel institutional structures have been developed to overcome particular constraints. Community colleges, for instance, were created in an attempt to democratize education by offering instruction at a lower cost and by accommodating people with less flexible schedules \cite{goldrick2010challenges}. Distance learning programs intended to deliver education in remote parts of the world and for people who simply could not attend in-person classes. Course materials, including assessments, were delivered through mail (correspondence education), radio, television, and eventually the Internet, thereby addressing geographical and time related constraints of traditional instruction \cite{moore1996distance}.

The latest generation of online learning environments, characterized by massive open online courses (MOOCs), has pushed the boundary on the scale of education \cite{waldrop2013campus}. By design, MOOCs provide course materials to millions of people worldwide. This scale could be achieved by pre-recording lectures, designing assessments that can be graded automatically, and by leveraging the momentum of the number of people involved (e.g., to facilitate peer learning or peer grading \cite{kulkarni2013peer,cambre2014talkabout}). Maybe by virtue of their large scale, their prominent instructors, or their adherence to contemporary interface designs, MOOCs rapidly became an online media phenomenon. People would sign up weeks in advance of the course launch date, many of whom would never even enter the course site. And among those who enroll and enter the site, a large proportion tends to only ``sample'' some content and leave again \cite{kizilcec2013deconstructing}. Many of the prototypical behaviors observed in MOOCs \cite{kizilcec2013deconstructing,breslow2013studying} resemble those on online media platforms. This trend has also been reflected in the diversity of MOOC learners' motivations for enrolling \cite{kizilcec2015motivation}.

Shortly after the first wave of courses had finished, extensive media coverage led to MOOCs becoming associated with high attrition rates \cite{lewin2013after,parr2013mooc,guthrie2013moocs}. Early MOOC research cautioned against dichotomizing learners into successes and failures based on course completion \cite{kizilcec2013deconstructing,rivard2013measuring}. Instead, more nuanced categorizations based on learner behavior \cite{kizilcec2013deconstructing,clow2013moocs}, motivations \cite{kizilcec2015motivation}, or intentions \cite{wilkowski2014student} have been suggested. Ultimately, perspectives on persistence and attrition in MOOCs depend on how MOOCs have been conceptualized. Kizilcec and Schneider \citeyear{kizilcec2015motivation} proposed that MOOCs have bridged two different world: one is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other adheres to the ``grammar of schooling'', which presupposes instructor-defined goals that students strive to achieve \cite{tyack1994grammar}. Viewing MOOC participation as bridging these two worlds undoubtedly adds a layer of complexity to interpretations of attrition. The present work extends the empirical base of research on attrition to inform the theoretical discourse and pave the path for novel practices.

This paper presents a systematic investigation of attrition in MOOCs, based on self-report and behavioral data collected from 20,879 online learners in 18 courses. A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their learning goals. We begin by briefly reviewing the large literature on attrition in educational environments with a focus on important developments in understanding its causes. Building on this foundation of prior work, Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions. In Study 2, to develop a deeper understanding of MOOC attrition, we invited around 6,500 learners who were predicted to have disengaged from a MOOC to provide feedback. Their open response answers about what challenges they faced while taking the MOOC were iteratively coded. We investigated the role of volitional control for learners who report having not enough time, and examine goal striving, social belonging, and growth mindset in relation to being successful in a MOOC.

@@

The majority of early work on attrition centered around theoretical models of students' decision to persist or dropout of a traditional higher education setting. The literature on attrition in higher education is largely concerned with students disengaging from a course of study rather than a single course. An early model suggested that students' presistence is largely driven by their prior behavior, attitudes, and norms \cite{fishbein1975belief}. The psychological processes involved in turning an intent to learn into the decision to persist were thought to be mediated by volition, i.e., the extent to which the student engages in goal-direct behaviors in the face of distraction \cite{corno1993role}. Hence, motivation alone is necessary but not sufficient for persistence. Students may fail to sustain efforts in the absence of strong self-regulatory skills.

\begin{description}
  \item[RQ1] How do prior learner characteristics (demographics, geographic location, intentions and motivations, and previous online courses) influence learners' likelihood of disengaging from a MOOC?
  \item[H1] Successful MOOC learners exhibit higher levels of goal striving than unsuccessful ones.
\end{description}

@@

\begin{description}
  \item[RQ2] What proportion of learners who disengaged were satisfied with their achievements in the course?
  \item[H3] Successful MOOC learners have a more fluid conception of intelligence (i.e., more of a growth than fixed mindset) than unsuccessful ones.
\end{description}

@@

\begin{description}
\item[RQ3a] What reasons do learners report for disengaging from a MOOC?
\item[RQ3b] How do reasons for disengaging vary across prior learner characteristics (cf. RQ1)?
\item[RQ4] How does behavior in MOOCs vary by reasons for disengaging?
\end{description}

@@

@@

% The two most commonly reported reasons for dropout are related to one's ability to commit time to the course. The correlation between the two reasons is $r=0.32$. A more detailed analysis we conditioned on responses to these two reasons revealed that 30\% of respondents who reported at least one of these two time reasons reported no other reasons for dropout, while 24\% reported one other reason and 46\% reported more than one. Difficulty of an assignment or exam was the third most reported reason for dropout. It is most positively correlated with reasons citing difficulties, including ``Too advanced for my background'' ($r=0.4$), ``Materials not explained properly'' ($r=0.37$), and ``Technical difficulties'' ($r=0.33$). The proportion of respondents who selected ``Technical difficulties'' exceeded 20\% in 10 out of the 17 courses. This reason, however, exhibits positive correlations stronger that 0.2 with 7 other reasons. [What might we like to study more about respondents who selected this reason?] In the median, 19\% of respondents reported they learned what they wanted in the course, and this reason was moderately positively correlated with reasons citing no or incomplete interest in the course, namely ``I was only exploring the course'' ($r=0.36$), ``The course is not challenging enough for me'' ($r=0.31$), and ``The course is not what I was looking for'' ($r=0.28$). Another set of interesting positive correlations exists between ``Course is not what I was looking for'' and other reasons citing the materials' level of difficulty/challenge, the value of the credentials to the respondent, and the extent to which she enjoyed the online course format, suggesting .... [TBC] The reason that exhibited the strongest positive correlations with other features was ``Videos were confusing''. it correlates most strongly with ``Materials not explained properly'' ($r=0.66$) and ``Did not enjoy online course format'' ($r=0.54$) Unlike correlations among reasons for disengagement, correlations of reasons for disengagement with enrollment intentions were mostly weak ($-0.14 < r < 0.24$). Among the strongest positive correlations were the cases listed in Table \ref{tab:rfd_ei_strongest_correlations}

@@

\section{Acknowledgments}
% We thank Elise Ogle and Ruth Bram for helping with the development of the codebook. 
% Thank Emily for feedback and Andreas for feedback and data support.
Omitted for blind review.
